### 2017-12-15 Writing/disparity_review.tex
rewritten intro (kindof)

Disparity, as a concept or set of toolkits is used a lot in biology and is more and more used every year.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Figures/GoogleScholarOccurences.pdf} 
\caption{Number of papers on Google Scholar matching the search ``morphological disparity'' per year. In black, the match is in the paper and in grey, in the title.}
\label{Fig:GoogleOccurences}
\end{figure}

In most of these papers, they trace back the disparity concept to seminal papers from the 90s: \cite{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution}.
However, being such a broad and obvious concept, it can probably be stretched back to way earlier.

@@

Biological data are complex; understanding the ecology and evolution of species often requires that we analyse multiple variables that covary with each other, and through space and time.
One solution to this problem is to analyse data in a multivariate framework.
Multivariate analyses aim to capture and incorporate this multidimensional complexity, while providing outputs that are interpretable in a physical world of only three dimensions.
Classical multidimensional species features that have been analysed through multivariate data ranges from their morphology \citep{raup1966geometric}, to their functional traits \citep{diaz2016global}.
However, such analysis are not limited at the species level and can also be applied to ecosystems \citep{DonohueDim}.

Key multidimensional features of species that have important roles in ecology and evolution include morphology, functional traits. etc. etc.
Many of these can be represented as matrices, these can be used in our package.
In the interests of clarity/brievety, we here focus on just one kind of data: morphological diversity.

Furthermore, we will also make a clear distinction between the multidimensional space which is a specific mathematical object and the disparity, which is a metric describing or summarising one or more aspects of this space.

Both can be defined in various ways depending on the authors, for example, disparity is defined as the weighted mean pairwise dissimilarity in \cite{Close2015} or the ellipsoid volume in \cite{DonohueDim} and the multidimensional space is defined as the morphospace in \cite{raup1966geometric} and the morpho-functional space in \cite{diaz2016global}.

The multidimensional space can be defined in many ways and arise from many mathematical transformations of the data such as the pairwise distance matrix \citep{Close2015}, a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, a multidimensional scaling \citep[MDS;][]{DonohueDim}, etc.
Similarly, disparity metrics \citep[or indices;][]{Hopkins2017} can defined in many ways \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015,diaz2016global}.
Finally, difference between disparity metrics can also be measured in many ways: using NPMANOVA \citep[e.g.][]{Brusatte12092008}, multidimensional permutation test \citep[e.g.][]{diaz2016global} or even simple confidence interval overlap \citep[e.g.][]{halliday2016eutherian}.

This variety of definitions and analysis have been developed in an equal variety of softwares such as \texttt{GINGKO} in javascript \citep{bouxin2005ginkgo,de2007ginkgo} or \texttt{geomorph} \citep{adams2013geomorph,adams2017geometric}, \texttt{Claddis} \citep{Claddis}, or \texttt{vegan} \citep{oksanen2007vegan} in R \citep{R}.
This results in the need to learn different languages (or at least - when restricted to R - different packages with different standards) as well as making analysis sometimes idiosyncratic and often complex to repeat since they are based on a particular feature from a particular software.
For example, in the excellent and widely used \texttt{geomorph} package morphological disparity analysis can be ran using the \texttt{morphol.disparity} function.
Unfortunately, however, the multidimensional space can only be defined as the ordination of the procrustes transform of geometric morphometric landmarks, the disparity can only be defined as the Procrustes variance and the difference between groups can only be measured through permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.

Disparity is an old concept blabalbal

\section{What \textit{is} disparity}

From a semantic view: morphospace or aspect of the morphospace?

From a biological view: what does an increase or a decrease in disparity represent?

\section{Why do we need disparity}

What are the fundamental questions we should use to answer disparity questions?

What are the questions we should not answer with disparity?

\section{What is missing}

Methods?
Data?

\section{Conclusion}

Yay!

% \section{Data availability and reproducibility}
% Data are available on Dryad or Figshare.
% Code for reproducing the analyses is available on GitHub (\url{github.com/TGuillerme/SpatioTemporal_Disparity}).

\section{Acknowledgments}
Royal Society..
I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.

>>>>

Biodiversity is not a smooth gradient of forms, functions, ecologies, behaviours, etc., but it is rather variable and discontinuous.
This is at the base of biological sciences and is linked to the concept of disparity \citep{Wills2001,Hopkins2017} with the observation that some groups of species are more similar or dissimilar than others.
This diversity in morphologies is in fact proper in nature and decoupled from taxonomic diversity: some groups may exhibit low diversity of species by high diversity of shapes \citep[or the other way around]{ruta2013,hopkinsdecoupling2013}.
In palaeobiology, this concept, has been studied since the 90s describing \citep{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution} and as been extended with modifications to macroevolutionary analysis in general since the 2000s (whether based on continuous traits; e.g. \citealt{Harmon961,geiger2008}; or on geometric morphometrics; e.g. \citealt{claude2008morphometrics,zelditch2012geometric,adams2013geomorph,adams2017geometric}).

Since then, morphological disparity has been used for testing a vast array of biological hypotheses such as: how did body plans evolve through time? \citep{Wesley-Hunt2005}; why does some morphologies exist and not other? \cite{gerber2017geometry}; what is the tempo and mode of morphological innovations through time? \citep{Hughes20082013}; what is the effect of mass extinction on disparity? \citep{halliday2016eutherian}; how does competition between groups affect disparity? \citep{Brusatte12092008}; how did morpho-ecological niches get filled through time? \citep{price2014niche}; and many more.
%
This apparent versatility of morphological disparity analysis led to a phenomenal amount of papers based on disparity methods (Fig. \ref{Fig:GoogleOccurences})!
%
However, ``a lot of disparity studies are inductive fishing trips and are not often designed to test a specific hypothesis.''
%
Indeed, compared to the vast amount of publication using morphological disparity, only a handful highlight/solve problems and propose hypothesis testing.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Figures/GoogleScholarOccurences.pdf} 
\caption{Number of papers on Google Scholar matching the search ``morphological disparity'' per year. In black, the match is in the paper and in grey, in the title. We collected the number of matches per year from 1960 to 2016 in Google Scholar for the terms "morphological disparity" both in the text (fuzzy matching) or in the title (exact matching). The data was collected on the 1st of November 2017.}
\label{Fig:GoogleOccurences}
\end{figure}

\section{What actually \textit{is} disparity?}

@@

Are all these methods the right tools to answer these questions (described in the intro) or are they meer “inductive fishing trips” and are actually not testing any of the hypothesis related to the disparity questions:

Can the data used in disparity studies actually test the phenomenon researchers are trying to capture?
E.g. for discrete morphological characters: disparity of fossilizable characters.
E.g. for morphometrics: disparity of the skull procrustes
E.g. for dtt: disparity of body/limb/beak length
Are  these data types enough to test disparity hypothesis or do we need more data/tone down the hypothesis?

Does the transformation of this data reflect our hypothesis? E.g. a morphospace as a pairwise matrix reflects the dissimilarity between the data, a morphospace based on an ordination reflects the variability within the data, a morphospace on both reflects the variability within the dissimilarity, are each method reflecting that?

Do the disparity metrics really reflect the variation in the hypothesis that are tested?
E.g. Are the variance/range based metrics a description of the ordination or the observed biological phenomenon (i.e. they are not including covariance and phylogenetic signal)
E.g. What does the average distance between species or from a point really convey? Is the distribution not more interesting?
Do we need more like metrics measuring explicitly what is tested in the hypothesis (rather than approximate it). For example, for competitive replacement, should we not measure whether group A occupies the same place as group B through time and whether that is due to competition or simply because any other place in the space cannot be occupied (and why are such spaces unoccupied)?

Are the methods used to test hypothesis in disparity actually testing what we think they’re testing (problem of multidimensional hypothesis)
What does an anova or related variance based test real infirms/confirms regarding the disparity questions?
How do we deal with bootstrapped data and pseudo-replications?
Can we realistically use univariate gaussian models to test hypothesis based on multivariate, probably non-gaussian data?

Conclusion:
We need a more hypothesis driven way of analysing disparity. Disparity methods should be the tool for answering disparity hypothesis, not merely a description of the data.

% \section{What \textit{is} disparity (really)?}

% Disparity can describe either the metric (disparity index \citep{Hopkins2017} or disparity metric \citep{Wills2001}) or the whole pipeline \citep{Claddis,zelditch2012geometric}

% % §Methods for data:
% \subsection{Data collection methodology}
% To assess the usage of disparity in different published studies, we collected methodological data from the 500 first Google Scholar results for the key words ``morphological disparity'' per order of appearance (accessed on the 1st of November 2017).
% For the 230 relevant papers among the 500 matches, we collected the following methodological data:

% \begin{itemize}
%     \item What was the focal biological group?
%     \item What kind of data was measured (e.g. landmarks, discrete data, etc.)?
%     \item Was data collected on the full organism or not?
%     \item How was the morphospace explicitly defined (e.g. PCA, PCO, MDS, etc.)?
%     \item How was the disparity metric(s) explicitly defined?
%     \item Which statistical test was applied to test the disparity related hypothesis?
%     \item Was phylogeny taken into account or not?
% \end{itemize}

% We used only the explicit definition of the morphospace and the disparity metric(s) in this search since a few number of papers had a vague definition of either or both (e.g. a disparity metric was measured but not described anywhere in the paper).

% The remaining 270 matches were disparity was mentioned but not measured felt in the following categories: papers out of topic, papers mentioning morphological disparity without measuring it, review papers, papers not accessible (either through a broken link or a paywall) or referenced citation without the paper (as a Google Scholar match).

% To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in a smaller number of categories (see supplementary materials and \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd}).

% \begin{figure}[!htbp]
% \centering
%    \includegraphics[width=1\textwidth]{Figures/MethodsProportions.pdf} 
% \caption{Disparity methods proportional usage: (1) Data type: which input data was used (...); (2) Morphospace: how was the morphospace obtained (...); (3) disparity: what type of disparity metric was calculated (...); (4) test: what type of test was applied (...).}
% \label{Fig:MethodsProportions}
% \end{figure}

% \hl{[ADD Biological group, full organism and phylogeny in the supplementary]}

% \noindent \hl{\textit{These points should be developed during the meeting}}

% \subsection{Disparity data}
% The data used for disparity methods comes from three main sources:
% \begin{itemize}
%     \item Continuous data such as limb or body measurements \citep{slaterCetacean}.
%     \item Discrete morphological characters \citep[sometimes referred to as ``Cladistic'' characters; ][]{Brusatte12092008}.
%     \item Geometric morphometric data which is generally the Procrustes transformation of 2D or 3D landmarks \citep{cooney2017mega}.
% \end{itemize}

% However, data was by far not limited to these three categories (e.g. colours \citealt{maia2013key}; metabolic rate \citealt{nespolo2016studying} or chemicals signal \citealt{garcia2017heterogeneous}).

% \subsection{Morphospaces}
% \begin{itemize}
%     \item PCAs [CITE].
%     \item PCOs [CITE]
%     \item Pairwise distance matrices \citep[no ordination; ][]{Harmon961}.
% \end{itemize}

% But also RWA [CITE] or Raup-spaces [CITE].

% \subsection{Disparity metrics}
% Throughout the 230 analysed papers, we found 103 unique combinations of metrics!

% \begin{itemize}
%     \item Distances measurements between species or from a certain point in the morphospace [CITE].
%     \item Ranges and variances of each axis of the morphospace \cite{Wills2001,Ciampaglio2001}
%     \item Distance between species based on the pairwise distance matrix (not the ordinations) [CITE].
% \end{itemize}

% But also metrics based on the volume [CITE], the characters dissimilarity [CITE] or the coordinates of some axis of the ordination [CITE]

% \subsection{Disparity hypothesis}
% Describing the many outputs (what and how is it tested):

% \begin{itemize}
%     \item Variance based (ANOVA, etc.) [CITE].
%     \item Correlation based [CITE].
%     \item Regression based (PGLS, etc.) [CITE].
% \end{itemize}

% But also discriminant analysis [CITE] and clustering [CITE].

% \subsection{The three main disparity analysis}
% From the collected data, we can highlight the use of three main different disparity analysis with their associated data/morphospace/metric/tests and related to specific methodological implementations.

% \begin{itemize}
%     \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data (sometimes referred to as ``Cladistic'' characters) for the full organisms to build a PCO from the organism's pairwise distance as a morphospace. Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g. the sum of variance or/and the sum of ranges). Hypothesis are often tested using multivariate ANOVAs on the pairwise distance matrix or by simply comparing the confidence intervals overlap of the disparity from different groups.

%     \item \textbf{The ``\texttt{geomorph}'' approach:} this group of method is based on landmark data (2D or 3D) on parts of the organism studied usually the skull) and use a Procrustes transformation of the landmarks that are then directly ordinated using a PCA (but sometimes RWA). Disparity is often measured as a distance metric (e.g. the distance between the species and a point in the morphospace such as its centroid). Hypothesis are then tested using ANOVA type tests with usually no phylogenetic correction (although phylogeny is sometimes used to correct the morphospace).

%     \item \textbf{The ``\texttt{dtt}'' approach:} this method can directly use continuous or discrete data for the full organism without any ordination (but not necessarily), and will measure disparity as the average pairwise distance between species (whether euclidean or any other type of distance). Hypothesis of higher/lower disparity can then be measured using null evolutionary models.

% \end{itemize}

% Of course some studies use a combination of these three methods or none of them at all!

% Also, among each category XX\% of studies use multiple approaches.

% [ADD Biological group, full organism and phylogeny in the supplementary]

% \section{Expanding disparity}

% How to compare disparity between groups? Is disparity relative?

% How to compare disparity between methods?

% Can we really say things about competition when looking at disparity in a single group?

% Are all characters equal? Character contingency suggests otherwise

% What is the relationship between disparity and tree shape?

% Do we have a null model for investigating disparity?

% Can we really say things about competition when looking at disparity in a single group?

% \noindent \hl{\textit{Disparity in other fields}}

% In ecology, disparity bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.

% \section{Conclusion}

% A quick guideline for good disparity analysis:

% Maybe we need something like in Parham et al 2012 (Best Practices for Justifying Fossil Calibrations): an easy an identifiable description of the pipeline containing: 1) the type of data, 2) the morphospace and 3) the metric?

\section{Acknowledgments}
The Royal Society.
TG acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.
